Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-04-27 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Add citations to the webarchive pages of the blog

This work has been done as an open collaboration by using \url{nmrlipids.blogspot.fi}
\todo{Current idea is that we will assign doi for the blog by using the help from  winnower administrators:
https://thewinnower.com/posts/archiving-and-aggregating-alternative-scholarly-content-dois-for-blogs.
Once done, that doi will be cited here and other places where blog content needs to be cited.} 
as a communication platform.
The approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
The present approach was started by publishing a manuscript~\cite{ollila13} discussing the glycerol backbone and choline structures 
in a Berger based model (the most used molecular dynamics simulation model for lipid bilayers).
After publishing the initial manuscript, the open invitation for further contributions and discussion through the blog was presented.
All the scientific contributions are done publicly through the blog and all contributors were offered coauthorship
according to the guidelines defined in the beginning of the project\todo{Citation to the On Credits in the blog here}. 
The acception of authorship is based on the self-assesment of the authors' scientific contribution to the project.
The contributions from each author are summarized in the Supplementary Information. %(the location depends on the journal probably).

@@

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_{CH}$ 
through scaling factor which depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{Citation to the blog will be added when available.}

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work as only $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters see \url{http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html}
\todo{Citation to the blog will be added when available.}

@@

In general there is a good agreement between the order parameters measured with different experimental NMR techniques: Almost all the 
reported values are within a variation of $\pm$0.02 (which is also the error estimate given by Gross et al.~\cite{gross97}) 
for all fully hydrated PC bilayer, regardless of the variation in their acyl chain composition and temperature.
Exceptions are the somewhat lower order parameters sometimes reported from measurements using $^1$H-$^{13}$C NMR~\cite{hong95a,hong95b,warschawski05}.
These experiments are not shown in Fig.~\ref{HGorderparameters} as the reported error bars are either relatively large~\cite{hong95a,hong95b}, 
or the spectral resolution is quite low and the numerical lineshape simulations have not been used in the analysis~\cite{warschawski05}.
Due to this end, it is highly likely that these reported lower order parameters are due to lower experimental 
accuracy and therefore we exclude them from our discussion. 
For more details, see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{Citation to the blog will be added when available.}
Motivated by the high experimental repeatability, we have highlighted in 
Fig.~\ref{HGorderparameters} the subjective sweet spots (light blue areas), within which we expect the calculated absolute 
values of the order parameters of a well-performing force field to fall.

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured (see
\url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
for more discussion.)
\todo{Citation to the blog will be added when available.}
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

@@

This work has been done as an open collaboration by using the \url{nmrlipids.blogspot.fi} as an communication
platform. All the scientific contributions have been done through the blog and are publicly
available. \todo{Citation to the blog will be added when available.}

@@

\todo{All authors, please add your acknowledgements.}

@@

\subsection{Author Contributions}
\todo{Please write a text summarizing your contribution} \\
{\it Alexandru Botan} provided simulation results for charmm36-UA\\
{\it Fernando Favela} prepared, performed and analyzed simulations
with popc and cholesterol for the CHARMM36 All-Atom FF. \\
{\it Patrick F. J. Fuchs} Ran and analyzed the Poger simulations. Provided scientific information which significantly advanced the project (signs and forking of order parameters). \\
{\it Matti Javanainen} prepared and performed simulations with multiple lipid models and analyzed the results. Supervised the work of JT.\\
{\it Matej Kanduc} provided simulation results for the Berger DLPC model \\
{\it Waldemar Kulig} prepared the MD simulations with cholesterol for the MacRog FF \\
{\it Antti Lamberg} Prepared and performed simulations with Berger and its variants to show the importance of the signs and stereospecific labeling of the order parameters. \\
{\it Claire Loison} provided simulation results for charmm36-UA. \\
{\it Alexander Lyubartsev} provided simulation results for Högberg08 force field.\\
{\it Markus S. Miettinen}  \\
{\it Luca Monticelli} Critical discussions in all phases of the project. Collaboration with Jukka  M{\"a}{\"a}tt{\"a}. \\
{\it Jukka M{\"a}{\"a}tt{\"a}}  prepared and performed simulations with Berger and Slipids models and analyzed the results.\\
{\it O. H. Samuli Ollila} Designed and managed the work. Ran and analyzed several simulations. Wrote the manuscript.  \\
{\it Marius Retegan} Prepared and validated the GROMACS compatible parameter files for GAFFlipid and Lipid14 force fields.\\
{\it Tomasz Rog} Provided the topologies for the MacRog model and also the full hydration simulation data with this model. \\
{\it Hubert Santuz} prepared and performed the cholesterol simulations with CHARMM36 and analyzed the results. \\
{\it Joona Tynkkynen} prepared and performed the dehydration simulations with the MacRog FF.\\

>>>>

This work has been done as an open collaboration by using \url{nmrlipids.blogspot.fi}~\cite{nmrlipids}
\todo{DONE}
as a communication platform.
The approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
The present approach was started by publishing a manuscript~\cite{ollila13} discussing the glycerol backbone and choline structures 
in a Berger based model (the most used molecular dynamics simulation model for lipid bilayers).
After publishing the initial manuscript, the open invitation for further contributions and discussion through the blog was presented.
All the scientific contributions are done publicly through the blog and all contributors were offered coauthorship
according to the guidelines defined in the beginning of the project~\cite{creditsPOST}\todo{DONE}. 
The acception of authorship is based on the self-assesment of the authors' scientific contribution to the project.
The contributions from each author are summarized in the Supplementary Information. %(the location depends on the journal probably).

@@

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_{CH}$ 
through scaling factor which depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see Ref.~\citenum{accuracyPOST}. \todo{DONE}

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work as only $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters see Ref.~\citenum{signpost}. \todo{DONE}

@@

In general there is a good agreement between the order parameters measured with different experimental NMR techniques: Almost all the 
reported values are within a variation of $\pm$0.02 (which is also the error estimate given by Gross et al.~\cite{gross97}) 
for all fully hydrated PC bilayer, regardless of the variation in their acyl chain composition and temperature.
Exceptions are the somewhat lower order parameters sometimes reported from measurements using $^1$H-$^{13}$C NMR~\cite{hong95a,hong95b,warschawski05}.
These experiments are not shown in Fig.~\ref{HGorderparameters} as the reported error bars are either relatively large~\cite{hong95a,hong95b}, 
or the spectral resolution is quite low and the numerical lineshape simulations have not been used in the analysis~\cite{warschawski05}.
Due to this end, it is highly likely that these reported lower order parameters are due to lower experimental 
accuracy and therefore we exclude them from our discussion. 
For more details, see Ref.~\citenum{accuracyPOST}. \todo{DONE}
Motivated by the high experimental repeatability, we have highlighted in 
Fig.~\ref{HGorderparameters} the subjective sweet spots (light blue areas), within which we expect the calculated absolute 
values of the order parameters of a well-performing force field to fall.

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion.) \todo{DONE}
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

@@

This work has been done as an open collaboration by using the \url{nmrlipids.blogspot.fi} as an communication
platform. All the scientific contributions have been done through the blog and are publicly
available~\cite{nmrlipids}. \todo{DONE}

@@

\todo{DONE}

@@

\subsection{Author Contributions}
{\it Alexandru Botan} provided simulation results for charmm36-UA\\
{\it Fernando Favela} prepared, performed and analyzed simulations
with popc and cholesterol for the CHARMM36 All-Atom FF. \\
{\it Patrick F. J. Fuchs} Ran and analyzed the Poger simulations. Provided scientific information which significantly advanced the project (signs and forking of order parameters). \\
{\it Matti Javanainen} prepared and performed simulations with multiple lipid models and analyzed the results. Supervised the work of JT.\\
{\it Matej Kanduc} provided simulation results for the Berger DLPC model \\
{\it Waldemar Kulig} prepared the MD simulations with cholesterol for the MacRog FF \\
{\it Antti Lamberg} Prepared and performed simulations with Berger and its variants to show the importance of the signs and stereospecific labeling of the order parameters. \\
{\it Claire Loison} provided simulation results for charmm36-UA. \\
{\it Alexander Lyubartsev} provided simulation results for Högberg08 force field.\\
{\it Markus S. Miettinen} \todo{Please write a text summarizing your contribution} \\
{\it Luca Monticelli} Critical discussions in all phases of the project. Collaboration with Jukka  M{\"a}{\"a}tt{\"a}. \\
{\it Jukka M{\"a}{\"a}tt{\"a}}  prepared and performed simulations with Berger and Slipids models and analyzed the results.\\
{\it O. H. Samuli Ollila} Designed and managed the work. Ran and analyzed several simulations. Wrote the manuscript.  \\
{\it Marius Retegan} Prepared and validated the GROMACS compatible parameter files for GAFFlipid and Lipid14 force fields.\\
{\it Tomasz Rog} Provided the topologies for the MacRog model and also the full hydration simulation data with this model. \\
{\it Hubert Santuz} prepared and performed the cholesterol simulations with CHARMM36 and analyzed the results. \\
{\it Joona Tynkkynen} prepared and performed the dehydration simulations with the MacRog FF.\\
